---
title: "Exploratory Analysis"
author: "Parsa Khazaeepoul, Yuna, Grace, Jinwoo"
date: "11/19/2021"
output: html_document

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(knitr)
unemployment_data <- read.csv("https://raw.githubusercontent.com/info201b-2021-aut/final-project-ydodobara/main/unemployment_data_us.csv?token=AV5HNBNQS2G3EMLZ3X2UWQLBT3JP2")
unemployment_by_race <- unemployment_data %>%
  group_by(Date) %>%
  select(Date, White, Black, Asian, Hispanic)
#Chart 1: Visualization of Average Unemployment Rate by Race Per Year 

race_unemployment <- unemployment_data %>%
  select(Year, White, Black, Asian, Hispanic)

race_unemployment_table <- race_unemployment %>%
  group_by(Year) %>%
  summarize(White = mean(White, na.rm = TRUE), Black = mean(Black, na.rm = TRUE), Asian = mean(Asian, na.rm = TRUE), Hispanic = mean(Hispanic, na.rm = TRUE)) %>%
  select(Year, White, Black, Asian, Hispanic)

race_unemployment_table <- pivot_longer(race_unemployment_table, White : Hispanic, names_to = "Race", values_to = "unemployment_rate")

race_unemployment_plot <- ggplot(race_unemployment_table) +
  geom_line(mapping = aes(x = Year, y = unemployment_rate, group = Race, color = Race)) +
  labs(x = "Year",
       y = "Unemployment Rate (in %)",
       title = "Average Unemployment Rate By Race Per Year")
race_unemployment_plot


#Chart 2: Visualization of Average Unemployment Rate by Academic
degree_unemployment <- unemployment_data %>% 
  select(Year, Primary_School, High_School, Associates_Degree, Professional_Degree) 
View(degree_unemployment)
degree_unemployment_table <- degree_unemployment %>%
  group_by(Year) %>%
  summarize(Primary = mean(Primary_School, na.rm = TRUE), High = mean(High_School, na.rm = TRUE), Associates = mean(Associates_Degree, na.rm = TRUE),  Professional = mean(Professional_Degree, na.rm = TRUE)) %>%
  select(Year, Primary, High, Associates, Professional)
View(degree_unemployment_table)

degree <- c("Primary", "High", "Associates", "Professional")

degree_unemployment_table <- data.frame(
  group=degree,
  value=c(sum(degree_unemployment_table$Primary),sum(degree_unemployment_table$High),sum(degree_unemployment_table$Associates), sum(degree_unemployment_table$Professional))
)


degree_unemployment_table

degree_unemployment_table <- degree_unemployment_table %>% 
  arrange(desc(group)) %>%
  mutate(prop = value / sum(degree_unemployment_table$value) *100) %>%
  mutate(ypos = cumsum(prop)- 0.5*prop )
degree_unemployment_table

proportion <- degree_unemployment_table$prop

degree_unemployment_plot <- ggplot(degree_unemployment_table, aes(x="", y=prop, fill=degree)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() + 
  theme(legend.position="top") +
  geom_text(aes(y = ypos, label = prop, size = 5), color = "white", size=6) +
  scale_fill_brewer(palette="Set1") + labs(title = "Percentage of Unemployment by Degree (in %)")

degree_unemployment_plot

#Chart 3: Visualization of Average Unemployment Rate by Gender Per Year 
race_unemployment <- unemployment_data %>%
  select(Year, Men, Women)

gender_unemployment_table <- race_unemployment %>%
  group_by(Year) %>%
  summarize(Men = mean(Men, na.rm = TRUE), Women = mean(Women, na.rm = TRUE)) %>%
  select(Year, Men, Women)

gender_unemployment_table <- pivot_longer(gender_unemployment_table, Men : Women, names_to = "Gender", values_to = "unemployment_rate")

gender_unemployment_plot <- ggplot(gender_unemployment_table) +
  geom_point(mapping = aes(x = Year, y = unemployment_rate, group = Gender, color = Gender)) +
  labs(x = "Year",
       y = "Unemployment Rate (in %)",
       title = "Average Unemployment Rate By Gender Per Year")
race_unemployment_plot



```

### Introduction
We chose for our domain to be unemployment rates in the U.S. on a national level. We chose unemployment rates as we hope to uncover trends and analyze the data to understand it better. Our data set includes various metrics that allow us to track education levels and races unemployment rates through time to help us better represent how rates have changed through the years. 

### Summary
We were able to gather a variety of information from the analysis of our data, and it can be best summarized as follows. In the year 2020, the average rate of employment for White people was 3.4%. For Black people in the same year, it was 6.2%. For Asian people, it was 3.2%. For Hispanic people, it was 4.9%. The Year and Associated month with the highest unemployment rate for White people was April 2010.

### Table of Aggregated Data (Grouped by Race)
```{r echo=FALSE, include=TRUE}

kable(unemployment_by_race, caption = "Unemployment rates grouped by race:")
```


### Charts displaying information from data (Visualizations)
```{r show_figure, fig.width = 6, fig.height = 6, message=FALSE,warning=FALSE,echo=FALSE}
race_unemployment_plot
degree_unemployment_plot
gender_unemployment_plot
```

The purpose for including chart one was to visually depict the relations between race and unemployment rate. The insight gained from this was that certain races had higher rates of unemployment then other races (Notably, Black and Hispanic's had the highest unemployment rates).

The purpose for including chart two was to visually depict the relations between degree and unemployment rate. The insight gained from this was that certain degrees had lower rates of unemployment than other degrees (Notably, as education levels rose, unemployment rates dropped). To visualize the differnce, pie chart was introduced.

The purpose for including chart th    ree was to visually depict the relations between gender and unemployment rate. The insight gained from this was that certain genders had lower rates of unemployment then other genders (Notably, Men had a higher unemployment rate than women).
